---
title: >
  Using GeneTonic on the macrophage dataset (EGAS00001000563)
author:
- name: Federico Marini
  affiliation: 
  - &id1 Institute of Medical Biostatistics, Epidemiology and Informatics (IMBEI), Mainz<br>
  - &id2 Center for Thrombosis and Hemostasis (CTH), Mainz<br>
  email: marinif@uni-mainz.de
- name: Annekathrin Ludt
  affiliation: 
  - *id1
- name: Jan Linke
  affiliation: 
  - *id1
  - *id2
- name: Konstantin Strauch
  affiliation: 
  - *id1

date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('ideal')`"
output: 
  bookdown::html_document2:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: show
    code_download: true
editor_options: 
  chunk_output_type: console
link-citations: true
bibliography: "../genetonic_supplement.bib"
---

```{r setup, include=FALSE, cache=FALSE, eval = TRUE, echo = FALSE}
library("knitr")
opts_chunk$set(
  fig.align = "center",
  fig.show = "asis",
  eval = TRUE,
  fig.width = 10,
  fig.height = 7,
  tidy = FALSE,
  message = FALSE,
  warning = FALSE,
  size = "small",
  comment = "##",
  echo = TRUE,
  results = "markup"
)
options(replace.assign = TRUE, width = 80)
```

# About the data

- Anka: EGAS00001000563
- Anka: macrophage package
- Anka: [@Alasoo2018]
- Anka: link to paper

# Loading required packages

We load the packages required to perform all the analytic steps presented in this document.

```{r loadLibraries, results='hide'}
library("DESeq2")
library("topGO")
library("pheatmap")
library("org.Mm.eg.db")
library("pcaExplorer")
library("ideal")
library("DT")
library("GeneTonic")
```

# Data processing

- Anka: refer to macrophage package
- Anka: refer to our vignette as well if you want

## Exploratory data analyisis

We read in the dataset, apply the vst transformation for performing PCA and creating a heatmap of the sample to sample distances.  
We'll use some functions from the `pcaExplorer` package [@Marini2019].

```{r eda-macrophage}
Anka: 

read in
rlog/vst
pca
heatmap?
```

## Differential expression analysis

We set the False Discovery Rate to 0.01 and we run the `DESeq2` workflow, generating results and using the `apeglm` shrinkage estimator.  
We plot the results as an MA-plot and report them in a table, using the functions from the `ideal` package [@Marini2020].

```{r de-macrophage}
Anka: 
FDR <- 0.01
test with increased lfcThreshold, see vignette
maplot
DT datatable



```

## Functional enrichment analysis

We perform functional enrichment analysis, here using the `topGOtable` wrapper to the method implemented in the `topGO` package.

- Anka: see backbone from SFile2
- Anka: main change: organism is human

```{r enrich-alma, cache=TRUE}
expressedInAssay <- (rowSums(assay(dds_alma)) > 0)
geneUniverseExprENS <- rownames(dds_alma)[expressedInAssay]
geneUniverseExpr <- anno_df$gene_name[match(geneUniverseExprENS, anno_df$gene_id)]



res_enrich <- shake_topGOtableResult(GObps_ko_vs_ctrl)
res_enrich <- get_aggrscores(
  res_enrich = res_enrich,
  res_de = res_alma_ko_vs_ctrl,
  annotation_obj = anno_df
)

# saveRDS(res_enrich, file = "usecase_res_enrich_alma.rds")

DT::datatable(res_enrich, rownames = FALSE)
```

### Using alternative enrichment analysis methods

It is possible to use the output of different other methods for enrichment analysis, thanks to the `shaker_*` functions implemented in `GeneTonic` - see for example `?shake_topGOtableResult`, and navigate to the manual pages for the other functions listed in the "See Also" section.

`GeneTonic` is able to convert for you the output of

- DAVID (text file, downloaded from the DAVID website)
- clusterProfiler (takes `enrichResult` objects)
- enrichr (a data.frame with the output of `enrichr`, or the text file exported from Enrichr)
- fgsea (the output of `fgsea()` function)
- g:Profiler (the text file output as exported from g:Profiler, or a data.frame with the output of `gost()` in `gprofiler2`)

Some examples on how to use them are reported here:

- Anka: "just port the macrophage part over?" + careful naming

```{r, eval=FALSE}
# clusterProfiler --------------------------------------------------------------
library("clusterProfiler")
degenes_alma <- deseqresult2df(res_alma_ko_vs_ctrl, FDR = 0.05)$id
ego_alma <- enrichGO(gene          = degenes_alma,
                     universe      = rownames(dds_alma)[expressedInAssay],
                     OrgDb         = org.Mm.eg.db,
                     ont           = "BP",
                     keyType       = 'ENSEMBL',
                     pAdjustMethod = "BH",
                     pvalueCutoff  = 0.01,
                     qvalueCutoff  = 0.05,
                     readable      = TRUE)
res_enrich_clusterprofiler <- shake_enrichResult(ego_alma)

# g:profiler -------------------------------------------------------------------
library("gprofiler2")
degenes <- deseqresult2df(res_alma_ko_vs_ctrl, FDR = 0.05)$SYMBOL
gostres_a20 <- gost(
  query = degenes, 
  ordered_query = FALSE, 
  multi_query = FALSE, 
  significant = FALSE, 
  exclude_iea = TRUE, 
  measure_underrepresentation = FALSE, 
  evcodes = TRUE, 
  user_threshold = 0.05, 
  correction_method = "g_SCS", 
  domain_scope = "annotated", 
  numeric_ns = "", 
  sources = "GO:BP", 
  as_short_link = FALSE)

res_enrich_gprofiler <- shake_gprofilerResult(gprofiler_output = gostres_a20$result)

# fgsea ------------------------------------------------------------------------
library("dplyr")
library("tibble")
library("fgsea")
library("msigdbr")
res2 <- res_alma_ko_vs_ctrl %>%
  as.data.frame() %>% 
  dplyr::select(SYMBOL, log2FoldChange)
de_ranks <- deframe(res2)
head(de_ranks, 20)
msigdbr_df <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP")
msigdbr_list = split(x = msigdbr_df$gene_symbol, f = msigdbr_df$gs_name)
fgseaRes <- fgsea(pathways = msigdbr_list, 
                  stats = de_ranks, 
                  nperm=100000)
fgseaRes <- fgseaRes %>% 
  arrange(desc(NES))

res_enrich_fgsea <- shake_fgseaResult(fgsea_output = fgseaRes)

# enrichr ----------------------------------------------------------------------
library("enrichR")
dbs <- c("GO_Molecular_Function_2018",
         "GO_Cellular_Component_2018",
         "GO_Biological_Process_2018",
         "KEGG_2019_Human",
         "Reactome_2016",
         "WikiPathways_2019_Human")
degenes <- (deseqresult2df(res_alma_ko_vs_ctrl, FDR = 0.05)$SYMBOL)
enrichr_output_a20 <- enrichr(degenes, dbs)

res_enrich_enrichr_BPs <- shake_enrichrResult(
  enrichr_output = enrichr_output_a20$GO_Biological_Process_2018)
res_enrich_enrichr_KEGG <- shake_enrichrResult(
  enrichr_output = enrichr_output_a20$KEGG_2019_Human)
```

## Assembling the `gtl` object

To simplify the usage of the function from `GeneTonic`, we create a `GeneTonic_list` object

- Anka: as above, 

```{r}
gtl_macrophage <- GeneTonic_list(
  dds = dds_alma,
  res_de = res_alma_ko_vs_ctrl,
  res_enrich = res_enrich,
  annotation_obj = anno_df
)

# saveRDS(gtl_alma, "gtl_alma.rds")
```

# Running `GeneTonic` on the dataset

Let's load the `GeneTonicList` object, containing all the structured input needed.

Anka: 
- from here onwards, let's use gtl_macrophage and use the correct references to that also in text

```{r}
gtl_alma <- readRDS("gtl_alma.rds")
```

## `GeneTonic`, interactively

This command will launch the `GeneTonic` app:

```{r eval=FALSE}
GeneTonic(gtl = gtl_alma)
```

The exploration can be guided by launching the introductory tours for each section of the app, and finally one can generate a small report focused on the features and genesets of interest.

If some more custom visualizations are required, it is possible to export the objects in use from the app as a SummarizedExperiment object, to be then passed to the `iSEE` software [@Rue-Albrecht2018] for further exploration.

## Using `GeneTonic`'s functions in analysis reports 

Anka: 
- I'd say we take 5-6 plots, and for the rest we refer to the vignette

The functionality of `GeneTonic` can be used also as standalone functions, to be called for example in existing analysis reports in RMarkdown, or R scripts.  
In the following chunks, we show how it is possible to call some of the functions on the A20-deficient microglia set.

```{r}
em <- enrichment_map(gtl = gtl_alma, n_gs = 100)

library("visNetwork")
library("magrittr")
em %>%
  visIgraph() %>%
  visOptions(
    highlightNearest = list(
      enabled = TRUE,
      degree = 1,
      hover = TRUE
    ),
    nodesIdSelection = TRUE
  )
```



# Session information {-}

```{r}
sessionInfo()
```

# Bibliography {-}
